---
title: "LVEXdetector Tutorial: Fitting and Prediction for Bisulfite Counts and Illumina Beta Values"
output:
  html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{LVEXdetector Tutorial: Fitting and Prediction for Bisulfite Counts and Illumina Beta Values}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

`LVEXdetector` provides functions for fitting the LVE-X model to two common types of DNA methylation data:

1. **Bisulfite sequencing count data**, using:
   - `LVEX_fit_bisulfite_counts()`
   - `LVEX_predict_bisulfite_counts()`

2. **Illumina-array beta values**, using:
   - `LVEX_fit_illumina_beta()`
   - `LVEX_predict_illumina_beta()`

In this tutorial, we illustrate both workflows using simulated data.

```{r load-package}
library(LVEXdetector)
```

# Part 1: Bisulfite sequencing counts

## Simulate female and male count data

In this example, female samples follow the LVE-X mixture structure, while male samples follow a simpler Binomial model driven by \eqn{\nu}.

```{r simulate-bisulfite}
set.seed(20250813)

# ----------------------------------------------------------
# Female simulation: Dirichlet-Multinomial style construction
# ----------------------------------------------------------
simulate_female_counts <- function(N = 100, phi = 50, depth_draw = 1000) {
  z  <- rnorm(N)
  w  <- rnorm(N)

  nu <- plogis(1 + 0.5 * z)
  pi <- plogis(-1 + 1 * w)

  # LVE-X category probabilities
  q2 <- nu * (1 - pi)
  q1 <- nu * pi * (1 - nu) + (1 - nu) * (1 - pi)
  q0 <- 1 - q1 - q2

  # total read depth before down-sampling to 50 pseudo-trials
  size <- rpois(N, depth_draw)

  # Dirichlet-multinomial construction using Dirichlet + Multinomial
  mu <- cbind(q0, q1, q2)
  eps <- 1e-12
  mu  <- pmax(mu, eps)
  mu  <- mu / rowSums(mu)

  alpha <- mu * phi

  counts <- matrix(0L, nrow = N, ncol = 3)
  colnames(counts) <- c("count0", "count1", "count2")

  for (i in seq_len(N)) {
    p_i <- as.numeric(extraDistr::rdirichlet(1, alpha[i, ]))
    counts[i, ] <- as.vector(rmultinom(1, size = size[i], prob = p_i))
  }

  # total successes among 2*size latent positions
  y_total <- counts[, 2] + 2L * counts[, 3]

  # down-sample to 50 positions via hypergeometric sampling
  M_total <- 2L * size
  K_success <- y_total
  n_draw <- 50L

  y_obs <- mapply(function(K, M, n) {
    rhyper(1, m = K, n = M - K, k = n)
  }, K = K_success, M = M_total, n = n_draw)

  list(
    MC = as.integer(y_obs),
    UC = as.integer(n_draw - y_obs),
    z  = z,
    w  = w
  )
}

# ----------------------------------------------------------
# Male simulation: Binomial
# ----------------------------------------------------------
simulate_male_counts <- function(N = 100, size = 50) {
  z  <- rnorm(N)
  w  <- rnorm(N)   # generated for dimensional consistency only
  nu <- plogis(1 + 0.5 * z)

  y <- rbinom(n = N, size = size, prob = nu)

  list(
    MC = as.integer(y),
    UC = as.integer(size - y),
    z  = z,
    w  = w
  )
}
```

## Generate one example dataset and fit the model

```{r fit-bisulfite-one}
N_fem  <- 100
N_male <- 100

fem_dat  <- simulate_female_counts(N = N_fem, phi = 50)
male_dat <- simulate_male_counts(N = N_male, size = 50)

MC <- matrix(c(fem_dat$MC, male_dat$MC), ncol = 1)
UC <- matrix(c(fem_dat$UC, male_dat$UC), ncol = 1)

z <- data.frame(z1 = c(fem_dat$z, male_dat$z))
w <- data.frame(w1 = c(fem_dat$w, male_dat$w))

fem_idx <- c(rep(TRUE, N_fem), rep(FALSE, N_male))

fit_counts <- LVEX_fit_bisulfite_counts(
  i = 1,
  MC = MC,
  UC = UC,
  z = z,
  w = w,
  fem_idx = fem_idx,
  bootstrap_on_fail = TRUE,
  bootstrap_B = 100,
  bootstrap_seed = 1
)

fit_counts
```

## Inspect fitted quantities

```{r inspect-bisulfite-fit}
fit_counts$point_estimates
fit_counts$standard_errors
head(fit_counts$nu_hat)
head(fit_counts$pi_hat)
head(fit_counts$nu_ci)
head(fit_counts$pi_ci)
fit_counts$ci_method
```

## Predict on new covariate values

```{r predict-bisulfite}
new_z <- data.frame(z1 = c(1, 0))
new_w <- data.frame(w1 = c(0, 1))

pred_counts <- LVEX_predict_bisulfite_counts(
  fit = fit_counts,
  new_z = new_z,
  new_w = new_w
)

pred_counts
```

## Small simulation study for bisulfite counts

This chunk runs multiple simulated datasets and summarizes bias, coverage, average model-based standard errors, and empirical standard deviations.

```{r sim-bisulfite, eval=FALSE}
set.seed(20250813)

D <- 200
N_fem <- 100
N_male <- 100
phi_true <- 50

theta_true <- c(
  a0 = 1,
  a1 = 0.5,
  b0 = -1,
  b1 = 1,
  kappa = phi_true
)

param_names <- c("a0", "a1", "b0", "b1", "kappa")

est_mat <- matrix(NA_real_, nrow = length(param_names), ncol = D)
se_mat  <- matrix(NA_real_, nrow = length(param_names), ncol = D)

rownames(est_mat) <- param_names
rownames(se_mat)  <- param_names
colnames(est_mat) <- paste0("sim", seq_len(D))
colnames(se_mat)  <- paste0("sim", seq_len(D))

fit_list_counts <- vector("list", D)

for (i in seq_len(D)) {
  fem_dat  <- simulate_female_counts(N = N_fem, phi = phi_true)
  male_dat <- simulate_male_counts(N = N_male, size = 50)

  MC_i <- matrix(c(fem_dat$MC, male_dat$MC), ncol = 1)
  UC_i <- matrix(c(fem_dat$UC, male_dat$UC), ncol = 1)

  z_i <- data.frame(z1 = c(fem_dat$z, male_dat$z))
  w_i <- data.frame(w1 = c(fem_dat$w, male_dat$w))

  fit_i <- LVEX_fit_bisulfite_counts(
    i = 1,
    MC = MC_i,
    UC = UC_i,
    z = z_i,
    w = w_i,
    fem_idx = c(rep(TRUE, N_fem), rep(FALSE, N_male)),
    bootstrap_on_fail = TRUE,
    bootstrap_B = 100,
    bootstrap_seed = 1000 + i
  )

  fit_list_counts[[i]] <- fit_i

  est_vec <- fit_i$point_estimates[c("a_Intercept", "a_z1", "b_Intercept", "b_w1", "kappa")]
  names(est_vec) <- param_names
  est_mat[, i] <- est_vec

  if (!is.null(fit_i$standard_errors)) {
    se_vec <- fit_i$standard_errors[c("a_Intercept", "a_z1", "b_Intercept", "b_w1", "kappa")]
    names(se_vec) <- param_names
    se_mat[, i] <- se_vec
  }
}

# Bias
bias_list <- list(
  a0    = est_mat["a0", ]    - theta_true["a0"],
  a1    = est_mat["a1", ]    - theta_true["a1"],
  b0    = est_mat["b0", ]    - theta_true["b0"],
  b1    = est_mat["b1", ]    - theta_true["b1"],
  kappa = est_mat["kappa", ] - theta_true["kappa"]
)

boxplot(
  bias_list,
  las = 2,
  main = "Bias of Parameter Estimates (Bisulfite counts)",
  ylab = "Bias (Estimate - Truth)",
  col = "lightblue"
)
abline(h = 0, col = "red", lty = 2)

# Coverage
coverage_counts <- sapply(param_names, function(p) {
  est <- as.vector(est_mat[p, ])
  se  <- as.vector(se_mat[p, ])

  ok <- is.finite(est) & is.finite(se)
  if (!any(ok)) return(NA_real_)

  lower <- est[ok] - 1.96 * se[ok]
  upper <- est[ok] + 1.96 * se[ok]

  mean(theta_true[p] >= lower & theta_true[p] <= upper)
})

coverage_counts

# Average model-based SE
sapply(param_names, function(p) mean(se_mat[p, ], na.rm = TRUE))

# Empirical SD
sapply(param_names, function(p) sd(est_mat[p, ], na.rm = TRUE))
```

# Part 2: Illumina methylation beta values

## Simulate female and male beta values

Here, female samples follow a Beta model whose mean is determined by the LVE-X mixture mean, while male samples follow a Beta model with mean \eqn{\nu} only.

```{r simulate-illumina}
set.seed(20250813)

# ----------------------------------------------------------
# Female simulation: Beta with LVE-X mixture mean
# ----------------------------------------------------------
simulate_female_beta <- function(N = 100, phi = 10, eps = 1e-12) {
  z  <- rnorm(N)
  w  <- rnorm(N)

  nu <- plogis(1 + 1 * z)
  pi <- plogis(-1 + 0.5 * w)

  q2 <- nu * (1 - pi)
  q1 <- nu * pi * (1 - nu) + (1 - nu) * (1 - pi)
  mu <- q2 + 0.5 * q1

  mu <- pmin(pmax(mu, eps), 1 - eps)
  y  <- rbeta(n = N, shape1 = mu * phi, shape2 = (1 - mu) * phi)

  list(y = y, z = z, w = w)
}

# ----------------------------------------------------------
# Male simulation: Beta with mean nu
# ----------------------------------------------------------
simulate_male_beta <- function(N = 100, phi = 10, eps = 1e-12) {
  z  <- rnorm(N)
  w  <- rnorm(N)

  nu <- plogis(1 + 1 * z)
  mu <- pmin(pmax(nu, eps), 1 - eps)

  y <- rbeta(n = N, shape1 = mu * phi, shape2 = (1 - mu) * phi)

  list(y = y, z = z, w = w)
}
```

## Generate one example dataset and fit the model

```{r fit-illumina-one}
N_fem  <- 100
N_male <- 100
phi_beta <- 10

fem_beta  <- simulate_female_beta(N = N_fem, phi = phi_beta)
male_beta <- simulate_male_beta(N = N_male, phi = phi_beta)

BetaMat <- matrix(c(fem_beta$y, male_beta$y), ncol = 1)

z_beta <- data.frame(z1 = c(fem_beta$z, male_beta$z))
w_beta <- data.frame(w1 = c(fem_beta$w, male_beta$w))

fem_idx_beta <- c(rep(TRUE, N_fem), rep(FALSE, N_male))

fit_beta <- LVEX_fit_illumina_beta(
  i = 1,
  BetaMat = BetaMat,
  z = z_beta,
  w = w_beta,
  fem_idx = fem_idx_beta,
  bootstrap_on_fail = TRUE,
  bootstrap_B = 100,
  bootstrap_seed = 1
)

fit_beta
```

## Inspect fitted quantities

```{r inspect-illumina-fit}
fit_beta$point_estimates
fit_beta$standard_errors
head(fit_beta$nu_hat)
head(fit_beta$pi_hat)
head(fit_beta$nu_ci)
head(fit_beta$pi_ci)
fit_beta$ci_method
```

## Predict on new covariate values

```{r predict-illumina}
new_z_beta <- data.frame(z1 = c(1, 0))
new_w_beta <- data.frame(w1 = c(0, 1))

pred_beta <- LVEX_predict_illumina_beta(
  fit = fit_beta,
  new_z = new_z_beta,
  new_w = new_w_beta
)

pred_beta
```

## Small simulation study for Illumina beta values

```{r sim-illumina, eval=FALSE}
set.seed(20250813)

D <- 200
N_fem <- 100
N_male <- 100
phi_true_beta <- 10

theta_true_beta <- c(
  a0 = 1,
  a1 = 1,
  b0 = -1,
  b1 = 0.5,
  kappa = phi_true_beta
)

param_names <- c("a0", "a1", "b0", "b1", "kappa")

est_mat_beta <- matrix(NA_real_, nrow = length(param_names), ncol = D)
se_mat_beta  <- matrix(NA_real_, nrow = length(param_names), ncol = D)

rownames(est_mat_beta) <- param_names
rownames(se_mat_beta)  <- param_names
colnames(est_mat_beta) <- paste0("sim", seq_len(D))
colnames(se_mat_beta)  <- paste0("sim", seq_len(D))

fit_list_beta <- vector("list", D)

for (i in seq_len(D)) {
  fem_i  <- simulate_female_beta(N = N_fem, phi = phi_true_beta)
  male_i <- simulate_male_beta(N = N_male, phi = phi_true_beta)

  y_i <- c(fem_i$y, male_i$y)
  z_i <- c(fem_i$z, male_i$z)
  w_i <- c(fem_i$w, male_i$w)

  BetaMat_i <- matrix(y_i, ncol = 1)
  z_i_df <- data.frame(z1 = z_i)
  w_i_df <- data.frame(w1 = w_i)

  fit_i <- LVEX_fit_illumina_beta(
    i = 1,
    BetaMat = BetaMat_i,
    z = z_i_df,
    w = w_i_df,
    fem_idx = c(rep(TRUE, N_fem), rep(FALSE, N_male)),
    bootstrap_on_fail = TRUE,
    bootstrap_B = 100,
    bootstrap_seed = 2000 + i
  )

  fit_list_beta[[i]] <- fit_i

  est_vec <- fit_i$point_estimates[c("a_Intercept", "a_z1", "b_Intercept", "b_w1", "kappa")]
  names(est_vec) <- param_names
  est_mat_beta[, i] <- est_vec

  if (!is.null(fit_i$standard_errors)) {
    se_vec <- fit_i$standard_errors[c("a_Intercept", "a_z1", "b_Intercept", "b_w1", "kappa")]
    names(se_vec) <- param_names
    se_mat_beta[, i] <- se_vec
  }
}

# Bias
bias_list_beta <- list(
  a0    = est_mat_beta["a0", ]    - theta_true_beta["a0"],
  a1    = est_mat_beta["a1", ]    - theta_true_beta["a1"],
  b0    = est_mat_beta["b0", ]    - theta_true_beta["b0"],
  b1    = est_mat_beta["b1", ]    - theta_true_beta["b1"],
  kappa = est_mat_beta["kappa", ] - theta_true_beta["kappa"]
)

boxplot(
  bias_list_beta,
  las = 2,
  main = "Bias of Parameter Estimates (Illumina beta values)",
  ylab = "Bias (Estimate - Truth)",
  col = "lightblue"
)
abline(h = 0, col = "red", lty = 2)

# Coverage
coverage_beta <- sapply(param_names, function(p) {
  est <- as.vector(est_mat_beta[p, ])
  se  <- as.vector(se_mat_beta[p, ])

  ok <- is.finite(est) & is.finite(se)
  if (!any(ok)) return(NA_real_)

  lower <- est[ok] - 1.96 * se[ok]
  upper <- est[ok] + 1.96 * se[ok]

  mean(theta_true_beta[p] >= lower & theta_true_beta[p] <= upper)
})

coverage_beta

# Average model-based SE
sapply(param_names, function(p) mean(se_mat_beta[p, ], na.rm = TRUE))

# Empirical SD
sapply(param_names, function(p) sd(est_mat_beta[p, ], na.rm = TRUE))
```

# Remarks

This tutorial showed how to:

- fit the LVE-X model to **bisulfite sequencing counts**,
- generate predictions for new covariate settings,
- fit the LVE-X model to **Illumina beta values**, and
- run simple repeated-simulation summaries for both settings.

In practice, you would replace the simulated inputs with your real methylation data matrices and sample-level covariates.
